---
title: "Asynchronous Longitudinal Regression"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Asynchronous Longitudinal Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup, message = FALSE, warning = FALSE}
library(skmle)
```

The two data frames come first in every call, so the estimators compose with
the native pipe. Examples below use it where it reads better.

## The problem

Suppose you want to know how a biomarker relates to a symptom score. The
biomarker is drawn at clinic visits; the symptom score is collected by a
questionnaire on its own schedule. Neither is measured often, and the two
schedules have nothing to do with each other. You have, per subject,

- a response process $Y_i(\cdot)$ observed at times $T_{i1}, \dots, T_{iL_i}$,
- a covariate process $X_i(\cdot)$ observed at times $S_{i1}, \dots, S_{iM_i}$,

and $\{T_{ij}\}$ and $\{S_{ik}\}$ never coincide. The model you want is the
ordinary one,

$$E\{Y_i(t) \mid X_i(t)\} = g\{X_i(t)^\top \beta\},$$

but $Y_i$ and $X_i$ are never observed at the same $t$, so not one row of the
data set is a complete observation of the regression you are trying to fit.

This is the setting of Cao, Zeng and Fine (2015). The package implements their
two estimators: `kee_async()` for a constant $\beta$, and `kee_async_td()` for
a coefficient curve $\beta(t)$.

## Two shortcuts that do not work

Before the estimator, it is worth being clear about why the obvious fixes fail,
because both are common in practice and both fail *silently* — they return
plausible numbers with plausible standard errors.

**Last value carried forward.** Replace $X_i(T_{ij})$ by the most recent
observed covariate value. This is measurement error with a twist: the error does
not shrink as the sample grows. With sparse sampling the expected gap back to
the previous observation is a property of the visit process, not of $n$, so the
substituted covariate keeps a fixed error variance forever. The result is
classical attenuation — $\hat\beta$ converges, but to something closer to zero
than $\beta$.

**Regression calibration.** Smooth the covariate path first, then substitute the
smoothed value. This feels more careful and has the same problem: to smooth at
$T_{ij}$ you need observations near $T_{ij}$, and under sparse sampling the
window cannot shrink without emptying. Whatever bandwidth keeps the window
occupied also keeps a non-vanishing lag spread, and the attenuation returns. For
a nonlinear link there is a second error on top: the estimating equation needs
the average of $g(\cdot)$, and you have supplied $g$ of the average.

What the kernel-weighted equation does instead is give up on reconstructing
$X_i(T_{ij})$ at all. It evaluates the link at each **observed** covariate
vector and lets a weight express how much that observation says about that
response occasion.

## The estimator

$$U_n(\beta) = n^{-1} \sum_i \sum_j \sum_k W_h(T_{ij} - S_{ik})\,
  X_i(S_{ik}) \left[ Y_i(T_{ij}) - g\{X_i(S_{ik})^\top \beta\} \right] = 0,$$

with $W_h(u) = W(u/h)/h$ and $W$ the Epanechnikov kernel. Every
(response, covariate) pair within a subject contributes, weighted by how far
apart in time the two occasions are. Pairs more than $h$ apart drop out; pairs
close together dominate.

Two things follow that are easy to miss.

First, **the rate is $(nh)^{1/2}$, not $\sqrt{n}$.** This is a smoothing
problem wearing a regression problem's clothes. Standard errors shrink more
slowly than you are used to, and the bandwidth is a real modelling choice rather
than a nuisance.

Second, **there is no requirement that time lie on $[0,1]$.** The equation
depends on times only through $(T - S)/h$, so any units work — provided `h` is
in the same units. That is the one thing to get right, and the package warns
when it looks wrong.

## A worked example

`sim_async_data()` generates the design from Section 4 of the paper: response
and covariate observation times are independent Poisson streams, and the
covariate is a Gaussian process.

```{r sim}
set.seed(2024)
d <- sim_async_data(n = 400, beta = c(0.5, 1.5))

str(d$y)
str(d$x)
```

Two tables, sharing only `id`. There is no way to merge them without inventing
rows, which is precisely the point.

```{r counts}
c(
  subjects = length(unique(d$y$id)),
  response_occasions = nrow(d$y),
  covariate_occasions = nrow(d$x),
  median_per_subject_y = median(table(d$y$id)),
  median_per_subject_x = median(table(d$x$id))
)
```

The formula spans both tables — left-hand side from `data_y`, right-hand side
from `data_x` — and `id` and `time` name columns present in each:

```{r fit}
fit <- d$y |>
  kee_async(d$x, y ~ x, id = id, time = time, h = 0.25)

summary(fit)
```

The truth is $\beta = (0.5, 1.5)$. `npair` records how many weighted pairs
actually contributed:

```{r npair}
c(pairs = fit$npair, subjects = nobs(fit))
```

Standard generics work as usual:

```{r generics}
coef(fit)
confint(fit)
sqrt(diag(vcov(fit)))
```

## Tidy output

`tidy()`, `glance()` and `augment()` return tibbles, so a fit goes straight into
the rest of a tidy workflow without reshaping.

```{r tidy}
tidy(fit, conf.int = TRUE)

glance(fit)
```

`augment()` attaches the fitted mean to the **covariate** table, because that is
where a fitted value lives here: the estimating equation evaluates the link at
each observed covariate vector, and it is the kernel weight, not the fitted
value, that connects it to a response occasion.

```{r augment}
augment(fit, d$x)
```

There is deliberately no `.resid` column. A residual needs a response value at
$S_{ik}$, which is exactly what asynchronous data does not have; reporting one
would mean inventing it. The fit does not retain its data, so `augment()` needs
`data_x` handed back to it.

## Choosing the bandwidth

Because the rate is $(nh)^{1/2}$, `h` trades variance against bias directly:
small `h` admits few pairs, large `h` admits pairs whose times are far apart and
so pulls in a bias. `kee_async_cv()` chooses it by cross-validation over
**subjects**, scoring each candidate by the kernel-weighted squared error on the
held-out subjects.

```{r cv}
cv <- d$y |>
  kee_async_cv(d$x, y ~ x,
    id = id, time = time,
    h_grid = c(0.05, 0.10, 0.15, 0.25, 0.40, 0.60),
    K = 5, seed = 1, quiet = TRUE
  )
cv
```

`cv_results` is a tibble, so it filters and plots directly:

```{r cv-results}
cv$cv_results
```

Two details of that criterion are worth knowing, because they determine whether
it is meaningful at all.

*Folds are over subjects, never rows.* Rows within a subject come from one
trajectory. Splitting by row would put the same subject on both sides and leak
the answer into the held-out score.

*The loss is divided by the total weight.* A wider bandwidth admits more pairs
and so accumulates more total squared error regardless of fit quality. Without
the denominator the criterion would select the smallest candidate every time,
which is not a bandwidth selection rule but a fixed point.

If the selected value lands on an endpoint of the grid, the function warns and
you should widen the grid: a minimum on the boundary is an artefact of where you
stopped looking.

Leaving `h_grid` unset generates one from the data, log-spaced over
$[2(Q_3 - Q_1)n^{-0.7},\, 2(Q_3 - Q_1)n^{-0.3}]$ where $Q_1, Q_3$ are quartiles
of the pooled observation times. It adapts to whatever units `time` is in.

## Sensitivity to the bandwidth is a diagnostic, not a nuisance

Cao, Zeng and Fine assume the covariance function of the covariate process is
twice differentiable, which gives a bias of order $h^2$. Their Section 6 notes
that relaxing this to one-sided differentiability — which admits processes with
independent increments, including the Ornstein-Uhlenbeck process that
`sim_async_data()` uses by default — leaves a bias of order $h$ instead.

The practical symptom is an estimate that drifts steadily as `h` grows. It costs
nothing to look:

```{r sweep, fig.height = 3.5}
hs <- c(0.05, 0.10, 0.15, 0.25, 0.40, 0.60)
sweep <- t(sapply(hs, function(h) {
  f <- d$y |> kee_async(d$x, y ~ x, id = id, time = time, h = h)
  c(h = h, est = coef(f)[2], se = sqrt(vcov(f)[2, 2]))
}))
round(sweep, 3)

plot(sweep[, "h"], sweep[, 2],
  type = "b", pch = 19, ylim = range(sweep[, 2] + c(-2, 2) * sweep[, 3]),
  xlab = "bandwidth h", ylab = expression(hat(beta)[1]),
  main = "Estimate against bandwidth"
)
arrows(sweep[, "h"], sweep[, 2] - 1.96 * sweep[, 3],
  sweep[, "h"], sweep[, 2] + 1.96 * sweep[, 3],
  angle = 90, code = 3, length = 0.04, col = "grey50"
)
abline(h = 1.5, lty = 2, col = "firebrick")
```

The estimate slides away from the truth as `h` grows, at a rate roughly linear
in `h`. That is the $O(h)$ term, and it is telling you that the covariance of
this covariate process has a kink at the diagonal. A plot that is flat across a
range of `h` says the opposite and is reassuring. Either way, report the plot,
not one bandwidth.

## Half kernel or full kernel

By default the kernel is two-sided: a response at $T_{ij}$ is informed by
covariate observations on either side of it. If the covariate must be *already
known* at the response time — a causal reading, or simply a design where later
covariate values could not have influenced the response — set
`one_sided = TRUE`.

```{r half}
full <- kee_async(d$y, d$x, y ~ x, id = id, time = time, h = 0.25)
half <- kee_async(d$y, d$x,
  y ~ x, id = id, time = time,
  h = 0.25, one_sided = TRUE
)

rbind(
  full = c(coef(full), pairs = full$npair),
  half = c(coef(half), pairs = half$npair)
)
```

The half kernel discards roughly half the pairs, so at a fixed `h` it is noisier.
Select `h` under the same setting you intend to fit with — `kee_async_cv()`
takes `one_sided` for exactly this reason.

## Time-varying coefficients

If the association itself changes over follow-up, `kee_async_td()` estimates
$\beta(t)$ pointwise. At a target time $t$ the response and covariate occasions
are weighted by their **separate** distances to $t$; the lag between them never
enters.

$$U_n\{\beta(t)\} = n^{-1} \sum_i \sum_j \sum_k W_{h_1}(t - T_{ij})\,
  W_{h_2}(t - S_{ik})\, X_i(S_{ik})
  \left[ Y_i(T_{ij}) - g\{X_i(S_{ik})^\top \beta(t)\} \right] = 0.$$

Because both time arguments are smoothed, the rate is the **bivariate**
$(n h_1 h_2)^{1/2}$. This is much slower than the time-invariant fit, and a
sample size that gave a comfortable constant-$\beta$ answer can be far too small
here.

```{r td, fig.height = 4}
set.seed(2025)
dt <- sim_async_data(
  n = 800, beta = function(tt) cbind(0.5, 1 + tt),
  lambda_y = 8, lambda_x = 8,
  x_cov = function(s, t) exp(-4 * (s - t)^2)
)

fit_td <- kee_async_td(dt$y, dt$x,
  y ~ x, id = id, time = time,
  times = seq(0.2, 0.8, by = 0.05), h = 0.2
)
plot(fit_td)
```

The truth is $\beta_1(t) = 1 + t$. Compare the fitted curve against it:

```{r td-truth}
tidy(fit_td, conf.int = TRUE) |>
  subset(term == "x", select = c(time, estimate, conf.low, conf.high)) |>
  head(4)
```

Three cautions when reading such a plot:

1. **The bands are pointwise, not simultaneous.** A curve leaving the band at
   one or two target times is not evidence of anything.
2. **They are not corrected for smoothing bias**, so they cover
   $E\hat\beta(t)$, not $\beta(t)$. A large bandwidth flattens the curve toward
   a constant and the bands do not widen to say so. Refit at a smaller `h` to
   see how much of the shape is smoothing.
3. **Edges are unreliable.** Target times within a bandwidth of the ends of the
   observed range draw on a one-sided window. Keep `times` inside the data.

If the bands cover a horizontal line across the whole range, the honest
conclusion is that the data do not resolve time variation — not that $\beta(t)$
is flat.

## Links other than the identity

`link = "log"` and `link = "logistic"` fit the corresponding generalised linear
models by Newton-Raphson on the same estimating equation. Everything above
applies unchanged; the cross-validation loss stays on the response scale.

```{r logistic}
set.seed(11)
db <- sim_async_data(n = 400, beta = c(0.3, 1.0), link = "logistic")
fit_b <- kee_async(db$y, db$x,
  y ~ x,
  id = id, time = time, h = 0.3, link = "logistic"
)
summary(fit_b)
```

## Getting the units right

The single most common mistake is a bandwidth on a different scale from the
observation times — times in days, `h` chosen as though they were on the unit
interval. Nothing about the estimator requires $[0,1]$, so the package cannot
simply reject the data; instead it tells you when almost nothing is
contributing.

```{r units, warning = TRUE}
d_days <- d
d_days$y$time <- d$y$time * 365
d_days$x$time <- d$x$time * 365

fit_days <- kee_async(d_days$y, d_days$x,
  y ~ x,
  id = id, time = time, h = 1
)
```

Rescaling `h` by the same factor recovers the original fit exactly, which is the
scale-equivariance of the estimating equation:

```{r units-fixed}
fit_ok <- kee_async(d_days$y, d_days$x,
  y ~ x,
  id = id, time = time, h = 0.25 * 365
)
all.equal(coef(fit_ok), coef(fit))
```

Note that this differs from the survival estimators in the package, which
*do* require times on $[0,1]$ because their spline basis and quadrature are
built there, and which stop with an error otherwise.

## Reference

Cao, H., Zeng, D. and Fine, J. P. (2015). Regression analysis of sparse
asynchronous longitudinal data. *Journal of the Royal Statistical Society,
Series B* **77**, 755–776.
