---
title: "Benchmarking skmle Against SurvSparse"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking skmle Against SurvSparse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

This vignette benchmarks `skmle` against the `SurvSparse` package on the same sparse longitudinal survival-data setting.

We compare:

1. `SurvSparse::add.haz()` against `skmle::kee_additive()` and `skmle::skmle(s = 1)`.
2. `SurvSparse::trans.haz()` against `skmle::skmle(s = 0)`.

The goal is not to claim a universal speed ratio for every problem size, but to show how the package behaves across a small grid of representative settings.

```{r setup, warning=FALSE, message=FALSE}
library(nloptr)
library(skmle)
library(SurvSparse)
library(survival)
library(dplyr)
library(bench)
```

## Data Generation

We simulate sparse longitudinal survival data using `sim_skmle_data()`. For a fair comparison with `SurvSparse`, which expects a single longitudinal covariate in these benchmark calls, we use the first simulated covariate.

```{r}
make_benchmark_data <- function(n, s_val) {
  dat <- sim_skmle_data(
    n = n,
    mu = function(tt) 8 * (0.75 + (0.5 - tt)^2),
    mu_bar = 8,
    alpha = function(tt) 0.5 * 0.75 + 0.75 * (tt * (1 - sin(2 * pi * (tt - 0.25)))),
    beta = c(1, -0.5),
    s = s_val,
    cen = 1.0
  )

  dat %>%
    dplyr::rename(covariates_old = covariates) %>%
    dplyr::mutate(
      covariates = covariates_old[, 1],
      X = X,
      delta = delta,
      obs_times = obs_times,
      id = as.integer(factor(id))
    ) %>%
    dplyr::select(id, X, covariates, obs_times, delta) %>%
    as.data.frame()
}
```

## Benchmark Helpers

```{r}
run_additive_benchmark <- function(n, iterations = 5) {
  dat <- make_benchmark_data(n, s_val = 1)
  h_val <- n^(-0.5)

  res <- bench::mark(
    SurvSparse_add_haz = add.haz(
      data = dat, n = n, tau = 1, h = h_val, method = 3
    ),
    skmle_kee_additive = kee_additive(
      Surv(X, delta) ~ covariates,
      data = dat, id = id, obs_times = obs_times, h = h_val
    ),
    skmle_spline = skmle(
      Surv(X, delta) ~ covariates,
      data = dat, id = id, obs_times = obs_times,
      s = 1, h = h_val, nknots = 3, norder = 3
    ),
    iterations = iterations,
    check = FALSE
  )

  as.data.frame(res[, c("expression", "median", "itr/sec")]) %>%
    dplyr::mutate(expression = as.character(expression)) %>%
    dplyr::mutate(
      scenario = sprintf("Additive (n = %d)", n),
      median_ms = as.numeric(median) / 1e6
    ) %>%
    dplyr::select(scenario, expression, median_ms, `itr/sec`)
}

run_transformed_benchmark <- function(n, iterations = 5) {
  dat <- make_benchmark_data(n, s_val = 0)
  h_val <- n^(-0.5)

  res <- bench::mark(
    SurvSparse_trans_haz = trans.haz(
      data = dat, n = n, nknots = 3, norder = 3, tau = 1, s = 0, h = h_val
    ),
    skmle_spline = skmle(
      Surv(X, delta) ~ covariates,
      data = dat, id = id, obs_times = obs_times,
      s = 0, h = h_val, nknots = 3, norder = 3
    ),
    iterations = iterations,
    check = FALSE
  )

  as.data.frame(res[, c("expression", "median", "itr/sec")]) %>%
    dplyr::mutate(expression = as.character(expression)) %>%
    dplyr::mutate(
      scenario = sprintf("Transformed (n = %d)", n),
      median_ms = as.numeric(median) / 1e6
    ) %>%
    dplyr::select(scenario, expression, median_ms, `itr/sec`)
}
```

## Results Across Multiple Scenarios

```{r benchmark-grid, warning=FALSE, message=FALSE}
set.seed(20260325)

benchmark_results <- dplyr::bind_rows(
  run_additive_benchmark(200, iterations = 5),
  run_additive_benchmark(500, iterations = 5),
  run_transformed_benchmark(100, iterations = 5),
  run_transformed_benchmark(200, iterations = 5)
)

benchmark_results
```

To make the speed comparison easier to read, the next table reports the ratio relative to the `skmle` method of interest in each scenario.

```{r speedup-table}
baseline_rows <- benchmark_results %>%
  dplyr::filter(
    (grepl("^Additive", scenario) & expression == "skmle_kee_additive") |
      (grepl("^Transformed", scenario) & expression == "skmle_spline")
  ) %>%
  dplyr::transmute(scenario, baseline_ms = median_ms)

speed_summary <- benchmark_results %>%
  dplyr::left_join(baseline_rows, by = "scenario") %>%
  dplyr::mutate(speedup_vs_baseline = median_ms / baseline_ms)

speed_summary
```

## Interpretation

The benchmark shows a stable pattern:

- For the additive comparison, `kee_additive()` is substantially faster than `SurvSparse::add.haz()`.
- For the transformed hazards comparison, `skmle(s = 0)` is much faster than `SurvSparse::trans.haz()` in these runs.
- The general spline-based `skmle(s = 1)` fit is slower than `kee_additive()`, which is expected because it solves the broader joint optimization problem rather than a specialized estimating equation.

In short, the package-level Rcpp implementation preserves the prototype method while moving the computational bottlenecks out of pure R.
