---
title: "Getting Started with skmle"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with skmle}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

`skmle` fits transformed hazards survival models when longitudinal covariates are observed sparsely and intermittently over time.

The package currently provides three main user-facing workflows:

1. `skmle()` for the general transformed hazards model.
2. `kee_cox()` for the proportional hazards estimating-equation approach.
3. `kee_additive()` for the additive hazards estimating-equation approach.

This vignette shows the usual modeling workflow: simulate data, fit a model, inspect the summary output, plot the estimated baseline component, and select a bandwidth by cross-validation.

```{r setup, warning=FALSE, message=FALSE}
library(nloptr)
library(skmle)
library(survival)
```

## Simulate Example Data

```{r}
set.seed(123)

dat <- sim_skmle_data(
  n = 80,
  mu = function(tt) 8 * (0.75 + (0.5 - tt)^2),
  mu_bar = 8,
  alpha = function(tt) 0.5 * 0.75 + 0.75 * (tt * (1 - sin(2 * pi * (tt - 0.25)))),
  beta = c(1, -0.5),
  s = 0,
  cen = 0.7
)

head(dat)
```

The simulated data are stored in long format. Each row corresponds to one observed longitudinal measurement time for one subject.

The key columns are:

- `id`: subject identifier
- `X`: observed event or censoring time
- `delta`: event indicator
- `covariates`: observed covariate values at that visit time
- `obs_times`: longitudinal observation time

## Fit the General Transformed Hazards Model

The `covariates` column returned by `sim_skmle_data()` is a matrix. You can either use it directly in the formula or split it into separate columns. Using the matrix directly is convenient for routine work.

```{r}
fit_skmle <- skmle(
  Surv(X, delta) ~ covariates,
  data = dat,
  id = id,
  obs_times = obs_times,
  s = 0,
  h = 0.5,
  nknots = 3,
  norder = 3
)

fit_skmle
```

The printed object gives the fitted coefficients. As in many R model objects, the formatted inferential output is produced by `summary()`.

```{r}
summary(fit_skmle)
```

The summary table reports:

- coefficient estimates
- standard errors
- z statistics
- p-values

## Plot the Estimated Baseline Component

```{r}
plot(fit_skmle)
```

This plot visualizes the estimated nonparametric baseline component from the sieve fit.

## Fit the Specialized Estimating-Equation Methods

When the model of interest matches one of the specialized settings, the package also provides dedicated estimating-equation estimators.

### Cox-Type Estimator

```{r}
fit_kee_cox <- kee_cox(
  Surv(X, delta) ~ covariates,
  data = dat,
  id = id,
  obs_times = obs_times,
  h = 0.5
)

summary(fit_kee_cox)
```

### Additive Hazards Estimator

For the additive hazards estimator, simulate data under `s = 1`.

```{r}
set.seed(456)

dat_add <- sim_skmle_data(
  n = 80,
  mu = function(tt) 8 * (0.75 + (0.5 - tt)^2),
  mu_bar = 8,
  alpha = function(tt) 0.75 + 0.75 * (tt * (1 - sin(2 * pi * (tt - 0.25)))),
  beta = c(1, -0.5),
  s = 1,
  cen = 0.7
)

fit_kee_add <- kee_additive(
  Surv(X, delta) ~ covariates,
  data = dat_add,
  id = id,
  obs_times = obs_times,
  h = 0.5
)

summary(fit_kee_add)
```

## Select a Bandwidth by Cross-Validation

Bandwidth selection can be handled by `skmle_cv()`.

```{r}
set.seed(999)

cv_fit <- skmle_cv(
  Surv(X, delta) ~ covariates,
  data = dat,
  id = id,
  obs_times = obs_times,
  s = 0,
  K = 3,
  h_grid = c(0.3, 0.4, 0.5),
  nknots = 3,
  norder = 3,
  quiet = TRUE
)

cv_fit$h_cv
cv_fit$cv_results
```

The returned object contains:

- the selected bandwidth
- the refitted `skmle` model at that bandwidth
- the cross-validation loss table

You can then inspect the final refit in the usual way.

```{r}
summary(cv_fit$fit)
```

## Typical Workflow

For routine use, the usual sequence is:

1. Prepare data in long format with one row per observation time.
2. Fit `skmle()` if you want the general transformed hazards model.
3. Use `summary()` and `plot()` to inspect the fitted model.
4. Use `skmle_cv()` if you want data-driven bandwidth selection.
5. Use `kee_cox()` or `kee_additive()` when the scientific model matches those specialized estimators.

This gives a standard R model-fitting interface while keeping the core numerical work in the Rcpp backend.
