---
title: "Getting Started with skmle"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with skmle}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

A longitudinal covariate is rarely measured at the time you need it. `skmle`
handles that mismatch by weighting each observation according to how far its
measurement time sits from the time being modelled, rather than carrying a value
forward or smoothing the covariate and substituting it.

Two outcome types are covered.

**Survival outcomes**, where the covariate is observed sparsely and
intermittently over follow-up:

1. `skmle()` for the general transformed hazards model.
2. `kee_cox()` for the proportional hazards estimating-equation approach.
3. `kee_additive()` for the additive hazards estimating-equation approach.

**Asynchronous longitudinal outcomes**, where the outcome is itself a sparsely
observed process recorded on a time grid that does not line up with the
covariate's:

4. `kee_async()` for time-invariant coefficients.
5. `kee_async_td()` for a coefficient curve $\beta(t)$.

This vignette walks through both: simulate data, fit a model, inspect the
summary output, plot the estimated baseline component, select a bandwidth by
cross-validation, then move to the asynchronous longitudinal setting.

## Which function do I need?

Two questions.

First, what is the outcome? A *time to an event* (death, relapse, failure),
possibly censored? Use the survival estimators; your data is one long table.
Or a *repeatedly measured quantity* (a score, a lab value) recorded on its own
schedule? Use the asynchronous estimators; your data is two tables.

Then:

| Outcome | Situation | Function |
| :-- | :-- | :-- |
| Survival | Start here; Cox model | `kee_cox()` |
| Survival | Additive hazards instead | `kee_additive()` |
| Survival | Want the baseline hazard, or a model between the two | `skmle()` |
| Survival | Choose the bandwidth properly | `skmle_cv()` |
| Longitudinal | Start here; one constant effect | `kee_async()` |
| Longitudinal | Choose the bandwidth properly | `kee_async_cv()` |
| Longitudinal | The effect may change over time | `kee_async_td()` |

Every fitting function picks a bandwidth for you if you do not supply one, and
says in a message what it chose. That is enough for a first answer; the `_cv()`
functions choose it from the data, which is what to report.

```{r setup, warning=FALSE, message=FALSE}
library(nloptr)
library(skmle)
library(survival)
```

## Simulating some data

```{r}
set.seed(123)

dat <- sim_skmle_data(
  n = 80,
  mu = function(tt) 8 * (0.75 + (0.5 - tt)^2),
  mu_bar = 8,
  alpha = function(tt) 0.5 * 0.75 + 0.75 * (tt * (1 - sin(2 * pi * (tt - 0.25)))),
  beta = c(1, -0.5),
  s = 0,
  cen = 0.7
)

head(dat)
```

The simulated data are stored in long format. Each row corresponds to one observed longitudinal measurement time for one subject.

The key columns are:

- `id`: subject identifier
- `X`: observed event or censoring time
- `delta`: event indicator
- `covariates`: observed covariate values at that visit time
- `obs_times`: longitudinal observation time

## Fitting the general transformed hazards model

The `covariates` column returned by `sim_skmle_data()` is a matrix. You can either use it directly in the formula or split it into separate columns. Using the matrix directly is convenient for routine work.

```{r}
fit_skmle <- skmle(
  Surv(X, delta) ~ covariates,
  data = dat,
  id = id,
  obs_times = obs_times,
  s = 0,
  h = 0.5,
  nknots = 3
)

fit_skmle
```

The printed object gives the fitted coefficients. As in many R model objects, the formatted inferential output is produced by `summary()`.

```{r}
summary(fit_skmle)
```

The summary table reports:

- coefficient estimates
- standard errors
- z statistics
- p-values

## Plotting the baseline

```{r}
plot(fit_skmle)
```

This plot visualizes the estimated nonparametric baseline component from the sieve fit.

## The specialised estimating equations

When the model of interest matches one of the specialized settings, the package also provides dedicated estimating-equation estimators.

### Cox-type

```{r}
fit_kee_cox <- kee_cox(
  Surv(X, delta) ~ covariates,
  data = dat,
  id = id,
  obs_times = obs_times,
  h = 0.5
)

summary(fit_kee_cox)
```

### Additive hazards

For the additive hazards estimator, simulate data under `s = 1`.

```{r}
set.seed(456)

dat_add <- sim_skmle_data(
  n = 80,
  mu = function(tt) 8 * (0.75 + (0.5 - tt)^2),
  mu_bar = 8,
  alpha = function(tt) 0.75 + 0.75 * (tt * (1 - sin(2 * pi * (tt - 0.25)))),
  beta = c(1, -0.5),
  s = 1,
  cen = 0.7
)

fit_kee_add <- kee_additive(
  Surv(X, delta) ~ covariates,
  data = dat_add,
  id = id,
  obs_times = obs_times,
  h = 0.5
)

summary(fit_kee_add)
```

## Selecting a bandwidth

Bandwidth selection can be handled by `skmle_cv()`.

```{r}
set.seed(999)

cv_fit <- skmle_cv(
  Surv(X, delta) ~ covariates,
  data = dat,
  id = id,
  obs_times = obs_times,
  s = 0,
  K = 3,
  h_grid = c(0.3, 0.4, 0.5),
  nknots = 3,
  quiet = TRUE
)

cv_fit$h_cv
cv_fit$cv_results
```

The returned object contains:

- the selected bandwidth
- the refitted `skmle` model at that bandwidth
- the cross-validation loss table

You can then inspect the final refit in the usual way.

```{r}
summary(cv_fit$fit)
```

## Half kernel or full kernel

Every estimator weights a covariate observation by how far it sits from the time
being modelled. By default that window is *one-sided*: only observations
strictly before the time contribute, which is the risk-set restriction of a
hazard model. Set `one_sided = FALSE` to smooth the covariate path from both
sides instead.

```{r one-sided}
fit_full <- kee_cox(
  Surv(X, delta) ~ covariates,
  data = dat, id = id, obs_times = obs_times,
  h = 0.5, one_sided = FALSE
)
cbind(half = coef(fit_kee_cox), full = coef(fit_full))
```

The switch reaches the risk-set averages inside the C++ backend as well as the
row weights, so the two halves of the estimator always use the same support.
`skmle_cv()` takes it too, so the bandwidth is selected under the same kernel
the final fit uses.

## Asynchronous longitudinal data

The estimators above model a survival outcome. When the outcome is itself a
sparsely observed longitudinal process, measured at times that do **not** line
up with the covariate's measurement times, use the kernel-weighted estimating
equations of Cao, Zeng and Fine (2015). Nothing is carried forward and no pair
is discarded: each (response, covariate) pair contributes in proportion to its
time separation.

```{r async-data}
set.seed(202)
d <- sim_async_data(n = 300, beta = c(0.5, 1.5))
head(d$y, 3)
head(d$x, 3)
```

The two tables are deliberately separate. They sit on different grids, so no
single data frame holds both without inventing rows. The formula
spans both: its left-hand side is looked up in `data_y`, its right-hand side in
`data_x`, and `id` and `time` name columns present in each.

```{r async-ti}
fit_a <- kee_async(d$y, d$x,
  y ~ x,
  id = id, time = time, h = 0.25
)
summary(fit_a)
confint(fit_a)
```

The bandwidth is the one consequential choice. `kee_async_cv()` selects it by
cross-validation over subjects, scoring each candidate by the kernel-weighted
squared error on the held-out subjects:

```{r async-cv}
cv <- kee_async_cv(d$y, d$x,
  y ~ x, id = id, time = time,
  h_grid = c(0.10, 0.15, 0.25, 0.40), K = 5, seed = 1, quiet = TRUE
)
cv
```

If the coefficients themselves vary with time, `kee_async_td()` estimates the
curve $\beta(t)$ pointwise. Both time arguments are smoothed there, so it
converges at the bivariate rate $(n h_1 h_2)^{1/2}$ and the bands are wide.

```{r async-td, fig.width = 7, fig.height = 4}
set.seed(203)
dt <- sim_async_data(
  n = 500, beta = function(tt) cbind(0.5, 1 + tt),
  lambda_y = 8, lambda_x = 8,
  x_cov = function(s, t) exp(-4 * (s - t)^2)
)
fit_td <- kee_async_td(dt$y, dt$x,
  y ~ x, id = id, time = time,
  times = seq(0.2, 0.8, by = 0.05), h = 0.2
)
plot(fit_td)
```

`vignette("asynchronous", package = "skmle")` covers this properly: why
last-value-carried-forward and regression calibration fail here, how to read
the bandwidth diagnostics, and what the half kernel changes.

## A typical workflow

For routine use, the usual sequence is:

1. Prepare data in long format with one row per observation time.
2. Fit `skmle()` if you want the general transformed hazards model.
3. Use `summary()` and `plot()` to inspect the fitted model.
4. Use `skmle_cv()` if you want data-driven bandwidth selection.
5. Use `kee_cox()` or `kee_additive()` when the scientific model matches those specialized estimators.
6. Use `kee_async()` or `kee_async_td()` when the outcome is longitudinal rather than a survival time.

This gives a standard R model-fitting interface while keeping the core numerical work in the Rcpp backend.
